
\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{tikz}
	\usetikzlibrary{calc}

\usepackage{FMC}

\newcommand{\commentout}[1]{}


\bibliographystyle{plainurl}


% - - - - - - - - - - - - - - -


%\newcommand\looppicx[1]{
%\begin{tikzpicture}[thick,x=12pt,y=12pt,outer sep=0pt,inner sep=2pt]
%	\coordinate (m) at (0,0);
%	
%	\begin{scope}[->,>=stealth,anchor=north]
%		\ifx#11
%		\draw (m)--node[pos=0.6,anchor=east]{$\termcolor\scriptstyle{i\vphantom j\,}$} (-1,-2) node {$\vphantom:\typecolor\scriptstyle{p}$};
%		\draw (m)--node[pos=0.6,anchor=west]{\,$\termcolor\scriptstyle{j}$}            ( 1,-2) node {$\vphantom:\typecolor\scriptstyle{q}$};
%		\else
%		\draw (m)--node[pos=0.6,anchor=east]{$\termcolor\scriptstyle{i\vphantom j\,}$} (-1,-2) node {$\vphantom:\typecolor\scriptstyle{\frac p{1-q}}$};
%		\fi
%	\end{scope}
%	
%	\node (x) at (0,-3) {};
%	
%	\coordinate (a) at ( .85,-1.7);% \node[gray,circle,draw,inner sep=0pt,minimum size=2pt] at (a) {};
%	\coordinate (b) at (1.25,-1.6);% \node[gray,circle,draw,inner sep=0pt,minimum size=2pt] at (b) {};
%	
%    	\ifx#10
%		%\draw (m) [rounded corners=20pt] -- (1.25,-2.5) [rounded corners=5pt] -- (1.25,1.5) -- (0,1.5) -- (0,1);
%		\draw[rounded corners=5pt] (m) -- (.5,-1) .. controls (a) and (b) .. (1.25,-.75) -- (1.25,1.5) -- (0,1.5) -- (0,1);
%    	\fi	
%	
%	\node[draw=red,fill=red!10,rounded corners,minimum size=16pt] (M) at (m) {$\term M$};
%	\draw[->,>=stealth] (0,2)--(M);	
%\end{tikzpicture}
%}

\newcommand\looppic[1]{
\vcenter{\hbox{%
\begin{tikzpicture}[thick,x=18pt,y=15pt,outer sep=0pt,inner sep=2pt]
	\coordinate (m) at (0,0);
	
	\node at (-0.3,-2) {$\dots$};
	
	\begin{scope}[->,>=stealth,anchor=north]
		\draw (m)--node[pos=0.7,anchor=east]{$\termcolor\scriptstyle{i_1\,}$} (-1,-2) node {$\vphantom:\typecolor\scriptstyle{\ifx#11p_1\else\frac{p_1}{1-q}\fi}$};
		\draw (m)--node[pos=0.7,anchor=east]{$\termcolor\scriptstyle{i_n\,}$} (.4,-2) node {$\vphantom:\typecolor\scriptstyle{\ifx#11p_n\else\frac{p_n}{1-q}\fi}$};
	\ifx#11
		\draw (m)--node[pos=0.7,anchor=west]{\,$\termcolor\scriptstyle{j}$}   ( 1,-2) node {$\vphantom:\typecolor\scriptstyle{q}$};
	\fi
	\end{scope}
	
	\node (x) at (0,-3) {};
	
	\coordinate (a) at ( .85,-1.7);% \node[gray,circle,draw,inner sep=0pt,minimum size=2pt] at (a) {};
	\coordinate (b) at (1.25,-1.6);% \node[gray,circle,draw,inner sep=0pt,minimum size=2pt] at (b) {};
	
    	\ifx#10
		%\draw (m) [rounded corners=20pt] -- (1.25,-2.5) [rounded corners=5pt] -- (1.25,1.5) -- (0,1.5) -- (0,1);
		\draw[rounded corners=5pt] (m) -- (.5,-1) .. controls (a) and (b) .. (1.25,-.75) -- (1.25,1.5) -- (0,1.5) -- (0,1);
    	\fi	
	
	\node[draw=red,fill=red!10,rounded corners,minimum size=16pt] (M) at (m) {$\term M$};
	\draw[->,>=stealth] (0,2)--(M);	
\end{tikzpicture}}}
}

\newcommand\markov[1]{\llparenthesis\kern1pt\term{#1}\kern1pt\rrparenthesis}
\newcommand\sem[1]{\llbracket#1\rrbracket}

\newcommand\PFMC{\textrm{PFMC}}
\newcommand\PFMCT{\PFMC$^{\,\smallbin{\Rightarrow\oplus}}$}


%==================================================================================================== FRONTMATTER

\title{Types for Almost-Sure Termination in the Functional Machine Calculus}

\author{Ugo {Dal Lago}}{University of Bologna, Italy \and INRIA Sophia Antipolis, France}{ugo.dallago@unibo.it}{https://orcid.org/0000-0001-9200-070X}{}

\author{Willem Heijltjes}{University of Bath, United Kingdom \and \url{http://willem.heijltj.es}}{w.b.heijltjes@bath.ac.uk}{https://orcid.org/0009-0001-8941-1150}{}

\authorrunning{U. Dal Lago and W. Heijltjes}

\Copyright{Ugo Dal Lago and Willem Heijltjes}

\ccsdesc[500]{Theory of computation~Lambda calculus}
\ccsdesc[300]{Theory of computation~Probabilistic computation}

\keywords{probabilistic lambda-calculus, type systems, almost-sure termination}

%\relatedversion{}

%\acknowledgements{Georgina Majury?}

%\nolinenumbers

\hideLIPIcs

\EventEditors{}
\EventNoEds{0}
\EventLongTitle{}
\EventShortTitle{}
\EventAcronym{}
\EventYear{}
\EventDate{}
\EventLocation{}
\EventLogo{}
\SeriesVolume{}
\ArticleNo{}

\newcommand{\UDL}[1]{\textcolor{brown}{Ugo: #1}}

%==================================================================================================== DOCUMENT
\begin{document}

\maketitle

\begin{abstract}
We investigate almost-sure termination in higher-order probabilistic computation through the lens of the functional machine calculus. Our key observation is that iteration  can easily be equipped with a non-zero chance of exiting, thus guaranteeing almost-surely terminating in a simple way. We exploit this insight by shifting attention from recursion to typed iteration, allowing probabilistic behavior to be tracked compositionally at the type level. For the purpose, we introduce a novel probabilistic type system that assigns explicit probability distributions to termination outcomes, and we show that these types soundly guarantee almost-sure termination, even in the presence of higher-order functions. We study the expressive power of the introduced system through an embedding of Markov chains, and prove that \emph{positive} almost-sure termination does not hold at higher-order, in general.
\end{abstract}




%---------------------------------------------------------------------------------------------------- INTRODUCTION

\section{Introduction}

We are interested in applying the toolset of the $\lambda$-calculus, in particular confluent reduction and type systems, to probabilistic computation. In~\cite{DalLago-Guerrieri-Heijltjes-2020} we restored confluence to a probabilistic $\lambda$-calculus, the probabilistic event $\lambda$-calculus (PEL), by decomposing probabilistic choice into a \emph{sampling} construct and a \emph{conditional}. Independently in~\cite{Antonelli-DalLago-Pistone-2022} and~\cite{Heijltjes-Majury-2025}, from different inspirations we converged on essentially the same probabilistic type system for the PEL to give lower bounds on the probability of termination. In this paper we further extend the calculus, and develop a type system for \emph{almost-sure termination} (AST): termination with probability one. In doing so, our objective differs from that of many existing works in this area, which aim to capture as many AST programs as possible~\cite{DalLago-Grellois-2019,DalLago-Faggian-Ronchi-2021}, or even to study the feasibility of capturing all such programs~\cite{Kobayashi-DalLago-Grellois-2020}. Instead, we seek to understand the extent to which almost-sure termination can be obtained and explained in a natural manner, possibly using well-established methods, and guaranteed by a minimal extension to simple types. More powerful methods such as dependent types and intersection types may later be used along standard lines to build on this.

We start from the following observation. \emph{Iteration}, generally represented by while-loops, repeats a given function $f$ until a condition is fulfilled. In the typed case, illustrated  below, the function $f$ is required to return a sum $B+A$ where the choice between returning a value $A$ or $B$ represents the condition. Then $\mathsf{iter}\,f$ evaluates $f$, exits on $B$, and repeats on $A$.
\[
\begin{array}{@{}r@{}l@{}}
	f &{}: A \to B+A
\\ \hline
	\mathsf{iter}\,f &{}: A\to B
\end{array}
\]
With the above rule, as with typed fixed-point combinators, typing no longer guarantees termination. However, we observe that if we replace $B+A$ by a probabilistic sum $B \oplus_p A$ with a fixed probability $p\in (0,1]$, i.e.\ the probability of choosing $B$ is non-zero and does not depend on the input, then $\mathsf{iter}\,f$ is \emph{almost-surely terminating} (provided of course that $f$ is itself terminating). This opens a way to approach almost-sure termination via type systems. A crucial question arises: how can we handle a type constructor such as $B\oplus_p A$ within a $\lambda$-calculus with probabilistic choice in a compositional and elegant manner?

These considerations illustrate how probabilistic choice is connected to other imperative features such as sequentiality and iteration. Derived from the PEL, the second author has developed a new approach to unifying the $\lambda$-calculus with imperative computation and effects, the \emph{Functional Machine Calculus} (FMC)~\cite{Heijltjes-2023}. Its most recent instance~\cite{Heijltjes-2025-MFPS} features branching sequential computation and iteration, providing an ideal vehicle for our present study. In the FMC each branch of the computation is labelled with a \emph{choice} label $i,j,k,\dots$, which takes the r\^ole of a constant, an exception, a case label, or an exit status. For example, the boolean constants $\bot$ and $\top$ are choice labels. Iteration of a term $\term M$ is written $\term{M^i}$, which evaluates $\term M$ and repeats if $\term M$ chooses $i$, terminating for any other choice. The calculus encodes a fair probabilistic choice, which we write as $\term{M+N}$.

\begin{example}
The term $\term{(\heads + (\tails + \,i))^i}$ chooses \emph{heads}, the choice label $\heads$, with probability $\frac12$, \emph{tails} $\tails$ with probability $\frac14$, and iterates with probability $\frac14$, to give an overall probability of $\frac 23$ for $\heads$ and $\frac 13$ for $\tails$. We illustrate it pictorially as follows.
\[
	\term{(\heads + (\tails + \,i))^i}
\quad=\quad
\vc{%
\begin{tikzpicture}[thick,x=20pt,y=24pt,outer sep=0pt,inner sep=1pt]
	\node (x) at (1,2) {$\termcolor\oplus$};
	\node (y) at (2,1) {$\termcolor\oplus$};

	\coordinate (a) at (2.75,0.25); %\node[gray,circle,draw,inner sep=0pt,minimum size=2pt] at (a) {};
	\coordinate (b) at (3   ,0.3);  %\node[gray,circle,draw,inner sep=0pt,minimum size=2pt] at (b) {};
	
	\begin{scope}[->,>=stealth,anchor=north]
	
	\draw (1,2.7)--(x);
	\draw (x)--(y);
	\draw (x) --node[pos=0.6,anchor=south east]{$\termcolor\scriptstyle{\heads}$} (0,1);
	\draw (y) --node[pos=0.6,anchor=south east]{$\termcolor\scriptstyle{\tails}$} (1,0);
	
	\draw[rounded corners=5pt] (y) -- node[pos=0.6,anchor=north east]{$\,\termcolor\scriptstyle{i}$}  
	  (2.5,0.5) .. controls (a) and (b) .. (3,0.7) -- (3,2.5) -- (1,2.5) -- (x);
	\end{scope}
\end{tikzpicture}}
\]
\end{example}

The type system we introduce in this paper, simplified for this introduction to omit inputs and return values, records the probability of terminating with each given choice. A term $\term M$ is typed as follows, indicating that it may choose $i_k$ with probability $p_k$ from choices $i_1$ through $i_n$.
\[
	\term{M : p_1 i_1 * \cdots * p_n i_n }
\]
The type is a probability (sub-)distribution over the choices $\{i_1,\dots,i_n\}$, written as a formal sum. For greater generality, and since our formalism readily incorporates it, we allow the probabilities $p_1$ through $p_n$ to sum to less than one, with the remainder representing an upper bound to the probability of divergence. We note here that, as a $\lambda$-calculus, the full calculus will have two means of non-termination: iteration with probability one, and higher-order non-terminating terms such as $\Omega=(\lambda x.xx)(\lambda x.xx)$. The type system will address both, and a term is almost-surely terminating if the sum probability of its type is exactly one.

\begin{example}
The body of our example term, without the loop, has the following type.
\[
	\term{\heads + (\tails + \,i) : \e ~=>~ /12\heads * /14\tails * /14i}
\]
\end{example}

Iteration $\term{M^i}$ removes the choice $i$ from the possible return choices. In the probabilistic setting, the remaining probabilities must then be renormalized to form a distribution. That is, if $i$ has probability $q$, the remaining choices (including divergence) have a total probability of $1-q$, and must be multiplied by $\frac1{1-q}$ to return the total to one. Typing for iteration is then as follows.
\[
\begin{array}{@{}r@{}l@{}}
	\term{M}   &{}: \type{p_1i_1 * \dots * p_ni_n * qj}
\\ \hline \\[-10pt]
	\term{M^j} &{}: \type{/{p_1}{1-q}i_1 * \dots * /{p_n}{1-q}i_n}
\end{array}	
\qquad
\term{M}=\looppic1 \qquad \term{M^j}=\looppic0
\]

\begin{example}
The return type for our example term is given by taking the distribution $\type{/12\heads * /14\tails * /14i}$, removing the element $\type{/14i}$, and renormalizing the remaining sub-distribution $\type{/12\heads * /14\tails}$ by multiplying by $\frac 43$, which yields the following.
\[
	\term{(\heads + (\tails + \,i))^i : /23\heads * /13\tails}
\]
\end{example}

Types for iteration then correctly give the probabilities for each choice. Consider $\term{M^j}$ where $j$ has probability $q$ in $\term M$. For any other choice $i$ with probability $p$ in $\term M$ the chance of $\term{M^j}$ choosing $i$ is given by the following geometric series, generated by choosing $i$ immediately, after one iteration, after two iterations, and so forth. The limit of this series is the probability given by the type system.
\[
	p + pq + pq^2 + pq^3 + \cdots ~=~ \frac p{1-q}
\]






The typed calculus achieves the following. A term is guaranteed to terminate with the probabilities of its return type, which is almost-sure termination if the sum probability is one, also in the presence of higher-order probabilistic functions. In a first-order setting it guarantees \emph{positive} almost-sure termination (expected evaluation time is finite), where still it may capture all Markov chains and directly compute their expected output. In the higher-order setting, which gives access to Church numerals, there are examples of \emph{non-positive} or \emph{null} AST.

The probabilistic type system is new for the FMC~\cite{Heijltjes-2025-MFPS}, while the underlying calculus remains essentially the same. It is also a conservative extension of the \emph{sequential} probabilistic event $\lambda$-calculus...

{\color{green}
TO DO:
\\~* decomposition of $N\oplus M$ into $\trm{?a.M}$ and $\trm{MaN}$
\\~* discuss full calculus (with stacks)
\\~* argue this is simple types (Remark~\ref{rem:simple types embed})
\\~* conservativity over previous work (Remark~\ref{rem:probabilistic types embed})
}

\commentout{\color{green!66!black}

* Conservative extension of [Curry \&\ Howard meet Borel] and [Simple Types for Probabilistic Termination].

* Uses \emph{iteration} rather than \emph{recursion} to obtain probabilistic typeability. Probabilistic recursion is difficult as it requires probabilistic function variables at type level, essentially using dependent types. Example: probabilistic Church numerals and Y-combinator.

* We believe the observation on typed iterators and almost-sure termination is new.

* Much work on probabilistic termination in verification, but not through type systems, which we believe is new.

* Limited expressive power for real algorithms at present, due to types being simple types. But a dependently typed generalisation would be akin to Hoare logic, and might facilitate probabilistic termination in theorem provers and give natural ways to include termination evidence in typed imperative programming languages.

}

%\cite{Hurd-2002}

%while (x>0) { x := x-1 [1/2] x := x+1 }

%  -1 (+) +1 : (=> p1.1 + p2.2 + p3.3 + ...) => (1/2 p1).0 + (1/2 p2).1 + (1/2 p1+p3).2 + (1/2 p2+p3).3 + ...


%---------------------------------------------------------------------------------------------------- BACKGROUND

\section{Related Work}

The type system introduced here can be seen as a conservative extension of the simple type systems studied in \cite{Antonelli-DalLago-Pistone-2022} and \cite{Heijltjes-Majury-2025}. While those systems already featured higher-order functions, they did not account for either iteration or recursion. In the present work, we deliberately choose \emph{iteration} rather than \emph{recursion} in order to keep the framework as simple as possible. Indeed, handling probabilistic recursion poses substantial technical challenges, as it typically requires probabilistic function variables at the type level, effectively leading to the use of dependent types.

Several other works \cite{Avanzini-DalLago-Ghyselen-2019,DalLago-Grellois-2019,DalLago-Padovani-2024,Das-Wang-Hoffmann-2023} introduce type systems that guarantee almost-sure termination (AST) or positive almost-sure termination (PAST) for probabilistic $\lambda$-calculi equipped with recursion. However, these approaches generally prioritise expressive power, often at the expense of conceptual and technical simplicity. A notable exception is the work of the first author with Breuvart and Herrou on AST for extensions of G\"odel’s $\mathbb{T}$~\cite{Breuvart-DalLago-Herrou-2021}. In that setting, however, one endows the underlying calculus either with recursion or with sampling from infinite distributions, rather than with iteration. In contrast, iteration is a fundamentally new ingredient of our approach and plays a central role in preserving simplicity while still capturing nontrivial probabilistic behaviour.

Another related line of research concerns the use of intersection types to characterise AST and PAST properties~\cite{DalLago-Faggian-Ronchi-2021}. Although such characterisations are indeed possible, they come at a significant cost: not only is type inference undecidable, but the resulting type structures are also highly complex. Yet another body of work focuses on the underlying verification problem. Unlike the deterministic case, probabilistic termination is undecidable even for higher-order programs operating over finite datatypes~\cite{Kobayashi-DalLago-Grellois-2020}, e.g., in probabilistic variants of the $\lambda Y$-calculus and higher-order recursion schemes. Our emphasis differs from these approaches, as we deliberately trade expressive power for a simpler and more tractable framework.

Finally, we acknowledge that the calculus introduced here currently has limited expressive power for practical algorithm design. This limitation stems both from the restriction to simple types and from the nature of exit probabilities in the iteration typing rule, which are fixed. A natural direction for future work would be a dependently typed generalisation, reminiscent of Hoare logic. Such an extension could facilitate reasoning about probabilistic termination in interactive theorem proving environments. From this perspective, the extensive body of work on sound verification of almost-sure termination in imperative programs~\cite{FerrerFioriti-Hermanns-2015,McIver-Morgan-Kaminski-Katoen-2018} and probabilistic automata~\cite{Etessami-Yannakakis-2009,Esparza-Kucera-Mayr-2004} also deserves mention. Although motivated by expressiveness and thus fundamentally different in spirit from the present work, this line of research offers many valuable ideas that could potentially be incorporated into the introduced framework.

%---------------------------------------------------------------------------------------------------- PROBABILISTIC FMC

\section{The Probabilistic FMC}

The Functional Machine Calculus (FMC)~\cite{Heijltjes-2023,Heijltjes-2025-MFPS} combines the $\lambda$-calculus with imperative programming and effects in the following way. Its point of departure is the (simplified) Krivine Abstract Machine~\cite{Krivine-2007}, a call--by--name operational semantics for the $\lambda$-calculus that evaluates terms in the context of an argument stack, where \emph{application} $M\,N$ pushes $N$, and \emph{abstraction} $\lambda x.M$ pops the head of the stack and substitutes it for $x$. The calculus and the machine are then extended as follows.

\begin{description}
	\item[Effects] The machine is generalised from one to multiple argument stacks or streams, which are then used to model effects: higher-order store as stacks of depth one, input/output as pop-only and push-only streams, and probabilistic choice as a pop-only stream of boolean values. The idea of a probabilistic $\lambda$-calculus that pops from a separate input stream derives from the probabilistic event $\lambda$-calculus~\cite{DalLago-Guerrieri-Heijltjes-2020}, the precursor to the FMC.

	\item[Sequencing] The calculus is extended with imperative \emph{skip} and \emph{sequence} $\term{M;N}$, implemented in standard fashion on the machine with a continuation stack where $\term{M;N}$ pushes $\term N$ and \emph{skip} pops and executes. This further enables the embedding of Plotkin's call--by--value $\lambda$-calculus, the monadic \emph{return} and \emph{let}-binding of Moggi's computational metalanguage~\cite{Moggi-1991}, and Levy's call--by--push--value~\cite{Levy-2003}.
	
	\item[Control] The calculus is generalised from sequential to branching computation by parameterizing \emph{skip} in a set of \emph{choice} labels $i,j,k,\dots$, each indicating a distinct branch of the computation. \emph{Sequence} becomes conditional on a choice $\term{M;i->N}$, composing $\term M$ with $\term N$ on the branch labelled $i$ in $\term M$, or operationally, evaluating $\term M$ and continuing with $\term N$ only if $\term M$ chooses the branch $i$. This is used to model control flow constructions: conditionals, constants, case switches, and exception handling. Iteration is introduced by the construct $\term{M^i}$ which evaluates $\term M$ and iterates on the choice $i$, exiting otherwise, with the semantics $\term{M^i}=\term{M;i->M^i}$.
\end{description}

\noindent
In this way the FMC embeds a minimal but complete imperative programming language, and extends to it two primary features of the $\lambda$-calculus: confluent reduction and simple types. In this paper we consider the \emph{probabilistic FMC} (\PFMC), a fragment of the calculus with \emph{control} and two argument stacks, the regular one for the $\lambda$-calculus and a probabilistic input stream. 

\begin{definition}
The \emph{terms} of the \PFMC\ are given by the following grammar.
\[
\term{M,N} ~\coloneqq~ \term{x} ~\mid~ \term{<x>.M} ~\mid~ \term{[N].M} ~\mid~ \term{i} ~\mid~ \term{M;i->N} ~\mid~ \term{M^i} ~\mid~ \term{?a.M}
\]
The constructs are: a \emph{variable} $\term x$; a \emph{pop} $\term{<x>.M}$ that binds $\term x$ in $\term M$; a \emph{push} $\term{[N].M}$; a \emph{choice} $\term i$; a \emph{case} $\term{M;i->N}$; a \emph{loop} $\term{M^i}$; and a \emph{generator} $\term{?a.M}$ which binds the variable $\term a$ in $\term M$.
\end{definition}

The $\lambda$-calculus embeds by letting $\lambda x.M=\term{<x>.M}$ and $M\,N=\term{[N].M}$. The generator $\term{?a.M}$ is rendered as $\term{`{\mathsf{rnd}}<a>.M}$ in other presentations of the FMC, as a parameterised pop construct accessing a stream $\mathsf{rnd}$ that represents a random generator. The calculus is thus an orthogonal combination of the (parameterized) $\lambda$-calculus, in the fragment of \emph{variable}, \emph{pop}, \emph{push}, and \emph{generator}, and the \emph{control} extension of the FMC, consisting of \emph{choice}, \emph{case}, and \emph{loop}.

Probabilistic choice $\term{M+N}$ is encoded as follows, where the \emph{case} construct associates left, $\term{M; i->N ; j->P}=\term{(M; i->N) ; j-> P}$.
\[
	\term{M+N} ~=~\term{?a.a ; \top->M ; \bot->N }
\]
In this construction, first $\term{?a}$ pops a boolean value $\term\top$ or $\term\bot$ from a probabilistic stream and substitutes it for $\term a$. The remainder then selects the relevant branch by the semantics of the \emph{case} construct, given by the equations $\term{i;i->M}=\term M$ and $\term{j;i->M}=\term j$ for $i\neq j$. We reserve the choice labels $\top$ and $\bot$ for the purpose of encoding probabilistic choice: here, if $\term M$ were allowed to choose $\term\bot$, that would incorrectly trigger $\term N$ after evaluating $\term M$.

\begin{example}
Our running example term unfolds into the full calculus as follows.
\[
	\term{(\heads+(\tails+\,i))^i}~=~\term{(?a.a ; \top->\heads ; \bot->(?b.b ; \top->\tails ; \bot -> i))^i}
\]
\end{example}

The generator $\term{?a.M}$ represents a \emph{fair} coin toss, with equal probability for $\term\top$ and $\term\bot$. We may write $\term{?a_p.M}$ for an unfair coin with probability $p$ for $\term\top$ and $1-p$ for $\term\bot$. This construct may either be considered a primitive, generalising $\term{?a.M}$, or constructed from fair coin tosses with iteration, in the way our running example term generates $\frac13$ probability.

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - Abstract machine
\subsection*{The Operational Semantics}

We will give the small-step and big-step operational semantics of the \PFMC, the former as an abstract machine and the latter as an inductive evaluation relation. The FMC further features a confluent reduction relation~\cite{Heijltjes-2025-MFPS}, which we omit here.

The abstract machine is given by the Krivine machine with the extensions described at the start of this section. In addition to the standard argument stack, it will have a \emph{trace}, a stack of boolean values that represents the probabilistic generator, and a \emph{continuation stack} that implements the branching and iteration of the \emph{case} and \emph{loop} constructs. A term $\term M$ is evaluated in the context of these three stacks until the machine halts with a choice term $\term i$ and an empty continuation stack, returning the argument stack as output and the choice $\term i$ as the exit status, indicating the chosen branch of the computation. The machine itself is deterministic; the probabilistic aspect of computation is modelled by taking the trace to be probabilistically generated.

%The Probabilistic FMC can be given meaning via an abstract machine whose states consist of the term under evaluation, two stacks (namely, the continuation stack and the evaluation stack) and a trace. The trace records the outcomes of future probabilistic choice instructions, thereby rendering the machine effectively deterministic. 
%More formally, the machine evaluates a term $\term M$ in a state $(t,S,\term M,K)$, in the context of three stacks:
%\begin{itemize}
%	\item A \emph{trace} $t$, a stack of boolean values $\top$ and $\bot$, from which $\term{?a.M}$ pops. 
%	\item An \emph{argument stack} $S$ of terms, where $\term{[N].M}$ pushes $\term N$, and $\term{<x>.M}$ pops the head $\term N$ and continues as $\term{\{N/x\}M}$, the capture-avoiding substitution of $\term N$ for $\term x$ in $\term M$.
%	\item A \emph{continuation stack} $K$, a stack of cases $\term{i->M}$, where $\term{M;i->N}$ pushes $\term{i->N}$ and $\term{M^i}$ pushes $\term{i->M^i}$. A choice $\term j$ pops the head $\term{i->M}$ and continues as $\term M$ if $i=j$, and as $\term j$ otherwise.
%\end{itemize}

\begin{definition}
The \emph{functional abstract machine} is given by the following data. \emph{Traces} $r,s,t$, \emph{argument stacks} $R,S,T$, and \emph{continuation stacks} $k,L$ are as follows.
\[
    r,s,t ~\coloneqq~ \e ~\mid~ \bot\,t ~\mid~ \top\,t   \qquad\quad 
    R,S,T ~\coloneqq~ \e ~\mid~ S\,\term M               \qquad\quad 
    K,L   ~\coloneqq~ \e ~\mid~ (\term{i->M})\,K
\]
The \emph{states} of the machine are four-tuples $(t,S,\term M,K)$. Its \emph{steps} or \emph{transitions} are given in Figure~\ref{fig:machine}. A \emph{run} of the machine is a sequence of steps, written with a double line as below. A \emph{successful} run is one terminating in a \emph{final state} $\state\e Ti\e$ with a choice term $\term i$ and empty trace and continuation stack.
\[ 
	\step= sSMK tTNL
\]
\end{definition}

\noindent
Example~\ref{example:run} in the appendix demonstrates a successful run of the machine for our running example term.

%The transitions of the machine implement the semantics as follows. The generator $\term{?a.M}$ pops $\top$ or $\bot$ from the trace $t$, 

%The continuation stack takes the form of a stack of cases $\term{i->M}$, where a term $\term{M;i->N}$ pushes $\term{i->N}$, a term $\term j$ pops $\term{i->M}$ and uses $\term M$ only if $i=j$, discarding it otherwise, and $\term{M^i}$ pushes $\term{i->M^i}$.

% - - - - - - - - - - Machine figure
\begin{figure}
\[
\begin{array}{lccr}
	  \mathsf{push}     & \step- t S {[N].M} K             t {S\,\term N} M K
	                    & \step- t S i {(\term{i->M})\,K}  t S M K                    & \mathsf{select}
\\ \\ \mathsf{pop}      & \step- t {S\,\term N} {<x>.M} K  t S {\{N/x\}M} K
	                    & \step- t S i {(\term{j->M})\,K}  t S i K                    & \mathsf{reject}
\\ \\ \mathsf{sequence} & \step- t S {M;i->N} K            t S M {(\term{i->N})\,K}
	                    & \step- t S {M^i} K               t S M {(\term{i->M^i})\,K} & \mathsf{iterate}
\\ \\ \mathsf{sample}   & \step- {\top\,t} S {?a.M} K      t S {\{\top/a\}M} K
                        & \step- {\bot\,t} S {?a.M} K      t S {\{\bot/a\}M} K
\end{array}
\]
\caption{The Abstract Machine}
\label{fig:machine}
\end{figure}
% - - - - - - - - - - 


The big-step operational semantics, in the form of an inductively defined evaluation relation $(\evalarrow)$, describes the successful runs of the machine directly. Where the machine implements branching with the contination stack, the big-step semantics gives a more abstract view as conditional sequential composition.

\begin{definition}
The \emph{evaluation} relation has statements $\eval tSM Ti$ and is defined inductively by the following rules.
\[
\begin{array}{ccc}
     \infer[\Rax]  {\eval \e Si Si}{}
   & \infer[\Rsel] {\eval {rs}R{M;i->N} Tj }{\eval rRM Si & \eval sSN Tj}
   & \infer[\Rrej] {\eval rR{M;i->N} Sj }{\eval rRM Sj}
\\\\ \infer[\Rpush]{\eval sS{[N].M} Ti}{\eval s{S\,\term N}M Ti}
   & \infer[\Rloop]{\eval {rs}R{M^i} Tj }{\eval rRM Si & \eval sS{M^i} Tj}
   & \infer[\Rexit]{\eval rR{M^i} Sj }{\eval rRM Sj} 
\\\\ \infer[\Rpop] {\eval t{S\,\term N}{<x>.M} Ti}{\eval tS{\{N/x\}M} Ti}
   & \infer[\Rbot] {\eval {\bot\,t}S{?a.M} Ti}{\eval tS{\{\bot/a\}M} Ti}
   & \infer[\Rtop] {\eval {\top\,t}S{?a.M} Ti}{\eval tS{\{\top/a\}M} Ti}
\end{array}
\]
\end{definition}

\noindent
The machine and the big-step evaluation relation can be put in correspondence as follows.

\begin{proposition}[{\cite[Proposition 5.4]{Heijltjes-2025-MFPS}}]
Evaluation characterizes successful machine termination:
\[
\eval tSMTi \quad \iff \quad \step= tSM\e \e Ti\e
\]
\end{proposition}


% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - Probabilistic evaluation

\subsection*{Probabilistic Evaluation}

The FMC as defined above is still deterministic: it externalises the probabilistic component of evaluation to the input trace $t$ and how this is generated. Here, we will define \emph{probabilistic} evaluation by considering distributions of machine runs over distributions of traces $t$. 

A \emph{frontier} $F$ is a (possibly infinite) set of traces such that no element in $F$ is a prefix of another. The \emph{probability} $\pr F$ of a frontier $F$ is the sum probability of its traces.
\[
\begin{array}{l@{}l}
	    \pr t &~=~ (\frac12)^{|t|} \quad\text{where $|t|$ is the length of $t$}
\\[5pt]	\pr F &~=~ \sum_{t\in F}\pr t
\end{array}
\]

\begin{example}
\label{ex:frontier}
The set of traces $\{ (\bot^n)\top\mid n\in \N \} = \{ \top,\bot\top,\bot\bot\top,\dots\}$ is a frontier with probability $1$.
\end{example}


A \emph{return family} $\DS$ gives the result of successfully evaluating a term for every trace in a given frontier $F$, collecting for each $t\in F$ a return stack $S_t$ and choice $i_t$ as the pair $(S_t,i_t)_t$ indexed with $t$. The frontier $F$ over which $\DS$ is defined is written $\floor\DS=F$, and the \emph{probability} of a return family is that of its frontier, $\pr\DS = \pr{\floor\DS}$.
\[
	\DS,\DT~\coloneqq~\{ (S_t,i_t)_t \}_{t\in F}
\]

\begin{definition}
A \emph{probabilistic evaluation} $\Eval SM\DT$ of a term $\term M$ in the context of a stack $S$ to a return family $\DT$ is given by the evaluation of $\term M$ and $\term S$ with each $t\in F$, as follows.
\[
	\Eval SM{\{ (T_t,i_t)_t \}_{t\in F}} \iff \forall t\in F.~(\eval tSM {T_t}{i_t})
\]
\end{definition}

\begin{example}
For the term $\term M=\term{(\heads + (\tails + \,i))^i}$ of the introduction we have $\Eval \e M \DT$ for the following return family $\DT$ over the frontier of Example~\ref{ex:frontier}.
\[
	\DT~=~\{ (\e,\heads)_t \mid t = (\bot^{2n})\top,~n\in\N~\} \cup \{ (\e,\tails)_t \mid t = (\bot^{2n+1})\top,~n\in\N~\}
\] 
\end{example}

Observe that evaluation is closed under inclusion: if $\Eval RM\DT$ and $\DS\subseteq\DT$ then $\Eval RM\DS$. An evaluation $\Eval SM\DT$ gives a lower bound of $\pr{\DT}$ for the probability of termination of a term $\term M$ in the context of a stack $S$.

\begin{definition}
The \emph{probability of termination} of a term $\term M$ in the context of a stack $S$ is the largest probability $\pr{\DT}$ for any $\DT$ such that $\Eval SM\DT$.
\end{definition}

To work with return families we use the following notation. The \emph{sum} of two return families $\DS+\DT$ is defined if they have disjoint frontiers, $\floor\DS\cap\floor\DT=\varnothing$, as the union of their underlying sets $\DS\cup\DT$. Likewise, \emph{inclusion} between families $\DS\subseteq\DT$ is inclusion as sets. We define the following notions below: $s\pfx F$ and $s\pfx\DS$ add $s$ as a prefix to every trace $t$ in $F$, respectively in $\DS$, and $S\,\DT$ prefixes every stack in $\DT$ with $S$; $\DS\only i$ restricts a family to the choice $i$ and $\DS\less i$ excludes the choice $i$; and $\stk\DS$ collects the stacks in $\DS$ as a set. 
\[
\begin{array}{r@{}l@{\qquad\qquad}r@{}l}
		s\pfx F    &{}~=~\{ \, st \mid t\in F \,\}                & \DS\only i &{}~=~\{ \,(S,j)_t \in \DS \mid j=i \,\} 
\\[5pt] s\pfx\DS   &{}~=~\{ \,(S,i)_{st} \mid (S,i)_t \in \DS\,\} & \DS\less i &{}~=~\{ \,(S,j)_t \in \DS \mid j\neq i \,\}
\\[5pt] S\,\DT     &{}~=~\{ \,(S\,T,i)_t \mid (T,i)_t \in \DT\,\} & \stk\DS    &{}~=~\{ \,S \mid (S,i)_t \in \DS\,\}
\end{array}
\]
We will freely use the following equations.
\[
	\floor{s\pfx\DS} = s\pfx\floor\DS    \qquad\quad
	\pr{s\pfx\DS}    = \pr s\times\pr\DS \qquad\quad
	\pr{\DS+\DT}     = \pr\DS + \pr\DT
\]


%---------------------------------------------------------------------------------------------------- TYPES

\section{Types}

Types for the FMC describe the input/output behaviour of the machine. In the deterministic case, a type $\type t$ is of the form below left, where the type vector $\type{!s}=\type{s_1 .. s_k}$ types the input stack on the machine, and the type vectors $\type{!t_m}$ type the return stack for each choice $i_m$. The type indicates that given an input stack $S:\type{!s}$ the machine will return some choice $i_m$ and stack $T:\type{!t_m}$. For the probabilistic case we replace the formal sum of the return type with a probability distribution, adding a probability $p_m$ for each summand as below right. For probabilities, we use rationals in $[0,1]$.
\[
	\type{!s ~=>~!t_1.i_1 + .. + !t_n.i_n} \qquad\qquad
	\type{!s ~=>~!t_1.p_1i_1 * .. * !t_n.p_ni_n}
\]
We will formalize the return type as \emph{distribution type}, a family in a finite set $I$ of choice labels, with a type vector $\type{!t_i}$ and a probability $p_i$ for each $i\in I$ such that the sum probability over all $p_i$ is at most one. It is thus a combination of a deterministic sum type, the family of type vectors $\type{!t_i}$, and a probability (sub-)distribution over $I$, the family of probabilities $p_i$.

\begin{definition}
\emph{Types}, \emph{stack types}, and \emph{distribution types} are given as follows.
\[
\begin{array}{l@{\qquad}r@{~\coloneqq~}l}
	\text{types}	            & \type{s,t} & \type{!t => d}
\\	\text{stack types}          & \type{!t}  & \type{t_1..t_n}
\\  \text{distribution types}   & \type{g,d} & \type{\{(!t_i,p_i)_i\}_{i\in I}}\quad\text{where}\quad\sum_{i\in I}p_i\leq 1
\end{array}
\]
\end{definition}

We introduce the following notation to work with distribution types. The \emph{basis} of a distribution type $\type d$, the set of choices $I$ over which it is defined, is $\floor{\type d}=I$. To recover the informal notation at the start of the section, we define the \emph{singleton} distribution type $\type{!t.pi}=\type{\{(!t,p)_i\}}$ and the \emph{sum} $\type{g*d}=\type{g\cup d}$ of distribution types over disjoint bases $\floor{\type g}\cap\floor{\type d}$. The \emph{probability} $\pr{\type d}$ of a distribution type is the sum over its probabilities $p_i$. Below, $\type{!s\,d}$ prefixes the type vector $\type{!s}$ to each $\type{!t}$ in $\type d$, and $\type{q.d}$ multiplies each probability $p_i$ in $\type d$ by the probability $q$. Below right, $\type{d|I}$ restricts $\type d$ to the choices in $I$, and $\type{d-I}$ removes them.
\[
\begin{array}{r@{}l@{\qquad\qquad}r@{}l}
	    \type{q.d}   &{}~=~ \type{\{    (!t,qp)_i \mid (!t,p)_i\in d\, \}}  & \type{d|I} &{}~=~ \type{\{ (!t.p)_i \in d \mid i\in I\,\}}
\\[5pt]	\type{!s\,d} &{}~=~ \type{\{ (!s\,!t,p)_i \mid (!t,p)_i\in d\, \}}  & \type{d-I} &{}~=~ \type{\{ (!t.p)_i \in d \mid i\notin I\,\}}
\end{array}
\]
Finally, the \emph{merger} $\type{g++d}$ of two distribution types is defined if for every $i\in\floor{\type g}\cap\floor{\type d}$ they have the same type vector $\type{!t_i}$, and is as follows.
\[
	\type{g++d} ~=~ \type{ \{ (!t,p+q)_i \mid i\in {\floor\gamma}\cup{\floor\delta}~`,~(!t,p)_i \in g~`,~(!t,q)_i\in d\}~+~g-{\floor\delta}~+~d-{\floor\gamma} }
\]

\begin{definition}
The \emph{typed probabilistic FMC} (\PFMCT\!) is given by the typing rules in Figure~\ref{fig:types}, where a context $\term{G}=\term{x_1:t_1,,x_n:t_n}$ is a finite function from variables to types written as a sequence, and typing judgment $\term{G |- M:t}$ assigns a term $\term M$ the type $\type t$ in the context $\term G$.
\end{definition}

% - - - - - - - - - - Types figure
\begin{figure}
\[
\begin{array}{l@{\qquad}c@{\qquad}c@{\qquad}r}
      \mathsf{variable}  & \vc{\infer{\term{G, x:t |- x:t}}{}}
                         & \vc{\infer{\term{G |- M : !s => 0 . d}}{}}                                 & \mathsf{zero}
\\ \\ \mathsf{pop}       & \vc{\infer{\term{G |- <x>.M : r\,!s => d}}{\term{G , x:r |- M: !s => d}}}
                         & \vc{\infer{\term{G |- [N].M : !s => d}}
                                     {\term{G |- N:r} && \term{G |- M : r\,!s => d}}}                 & \mathsf{push}
\\ \\ \mathsf{choice}    & \vc{\infer{\term{G |- i: !s => !s.1i}}{}}
                         & \vc{\infer{\term{G |- M;i->N : !r => g ++ (p . d)}}
                                     {\term{G |- M: !r => g * !s.pi} && 
                                      \term{G |- N: !s => d}}}                                        & \mathsf{sequence}
\\ \\ \mathsf{expand}    & \vc{\infer{\term{G |- M: !s\,!r => !r\,d}}{\term{G |- M: !s => d}}}
                         & \vc{\infer{\term{G |- M^i: !s => /1{1-p} . d}}
                                     {\term{G |- M: !s => d * !s.pi} && (p\neq 1)} }                  & \mathsf{iterate}    
\\ \\ \mathsf{sample}    & \multicolumn{2}{c}{
                           \vc{\infer{\term{G |- ?a.M : !r => (/12 . g) ++ (/12 . d)}}
                                     {\term{G , a: \e => \e.1\top |- M: !r => g} && 
                                      \term{G , a: \e => \e.1\bot |- M: !r => d}}}
                          }
\end{array}
\]
\caption{The typed probabilistic FMC}
\label{fig:types}
\end{figure}
% - - - - - - - - - -

\begin{example}
The probabilistic sum $\term{M+N}=\term{?a.a ; \top->M ; \bot->N }$ is typed as follows.
\[
\infer{ \term{|- ?a.a ; \top -> M ; \bot -> N : !t ~=>~ /12.g ~++~ /12.d} }{
 \infer{ \term{|- ?a.a ; \top -> M : !t ~=>~ !t./12\top ~*~ /12.d } }{
  \infer{ \term{|- ?a.a : !t ~=>~ !t./12\top ~*~ !t./12\bot} }{
   \infer{ \term{|- ?a.a : \e ~=>~ \e./12\top ~*~ \e./12\bot} }{
    \infer{ \term{ a: \e => \e.1\top |- a: \e => \e.1\top} }{}
    &&
    \infer{ \term{ a: \e => \e.1\bot |- a: \e => \e.1\bot} }{}
  }} &
  \term{|- M : !t => g}
 } &
 \term{|- N : !t => d}
}
\]
Using this as an admissible typing rule, our running example is typed as follows.
\[
\infer{ \term{|- (\heads + (\tails + \,i))^i : \e ~=>~ \e./23\heads ~*~ \e./13\tails} }{
 \infer{ \term{|- \heads + (\tails + \,i) : \e ~=>~ \e./12\heads ~*~ \e./14\tails ~*~ \e./14i} }{
  \infer{ \term{|- \heads : \e ~=>~\e.1\heads}}{}
  &&
  \infer{ \term{|- \tails + \,i : \e ~=>~ \e./12\tails ~*~ \e./12i } }{
   \infer{ \term{|- \tails : \e ~=>~\e.1\tails}}{}
   &&
   \infer{ \term{|- i : \e ~=>~\e.1i}}{}
}}}
\]
\end{example}

\begin{remark}
\label{rem:simple types embed}
The type system is a conservative extension of simple types for the $\lambda$-calculus. These embed as follows, using a single default choice $\star$ (see also~\cite{Heijltjes-2023,Heijltjes-Majury-2025,Heijltjes-2025-MFPS}). For these types, the typing rules for $\mathsf{variable}$, $\mathsf{pop}$, and $\mathsf{push}$ are those of the simply typed lambda-calculus.
\[
	\typ{t_1 \to \dots \to t_n \to o} \quad=\quad \type{t_1 \dots t_n => \e.1\star}
\]
\end{remark}

\begin{remark}
\label{rem:probabilistic types embed}
The type system is also a conservative extension of those of our previous work on probabilistic termination~\cite{Antonelli-DalLago-Pistone-2022,Heijltjes-Majury-2025}. Using the notation of~\cite{Heijltjes-Majury-2025}, simple probabilistic types are simple types over the rationals between $0$ and $1$ as the set of base types, and embed as follows, where $p\in\mathbb{Q}\cap[0,1]$ and $\star$ is again the single available choice label.
\[
	\typ{t_1 \to \dots \to t_n \to p} \quad=\quad \type{t_1 \dots t_n => \e.p\star}
\]
\emph{Sequential probabilistic types}~\cite{Heijltjes-Majury-2025} are the restriction of the present system to the choice $\star$.
\[
	\typ{!s} \overset{p}{\Rightarrow} \typ{!t} \quad=\quad \type{!s => !t.p\star}
\]
\end{remark}

\begin{remark}
For almost-sure termination, where the probabilities for a distribution type sum to one, types may be further simplified by omitting probabilities altogether, $\type{t}~\coloneqq~\type{!s => !t_1.i_1 + .. + !t_n.i_n}$, by the following reasoning. Firstly, the $\mathsf{zero}$ typing rule is invalid, and omitted. As can be observed from the remaining typing rules, all probabilities in a typed term are then non-zero. For the $\mathsf{iterate}$ rule with a premise $\term{M:!s => d * !s.pi}$, the condition $p\neq 1$, that the chance to iterate is less than one, is then automatically fulfilled if $\type d$ is non-empty. This is the only point where the type system places a condition on the probabilities; hence they may be omitted.
\end{remark}

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - Almost sure termination
\subsection*{Almost-Sure Termination}

The intended meaning of the type system is that if $\term{|- M: \e => d}$ (i.e.\ $\term M$ requires no inputs to run on the machine) then the probability of termination for $\term M$ is at least $\pr{\type d}$. We will establish this using a reducibility argument in the style of Tait~\cite{Tait-1967}, called \emph{runnability} since it uses the machine rather than reduction. 
The runnability predicate directly interprets the meaning of types: a runnable term of type $\term{M:!s => d}$, given a stack $S$ of runnable terms for the types in $\type{!s}$, will probabilistically evaluate to a runnable return family $\DT$ for the type $\type{d}$, with the key property that $\pr{\DT}\geq\pr{\type d}$.

\begin{definition}
The set $\RUN{t}$ of \emph{runnable terms} for a type $\type t$ is defined as the set of closed terms
\[
	\RUN{!s => d} = \{ \term M ~\mid~ \forall S \in\RUN{!s}.~\exists\DT\in\RUN{d}.~\Eval SM\DT~\}
\]
where the runnable sets for stack types $\type{!t}$ and distribution types $\type{d}$ are as follows.
\[
    \RUN{t_1..t_n} = \{ \e\,\term{M_1}\dots \term{M_n} \mid \term{M_i}\in\RUN{t_i} \}
\]
\[
	\RUN{\{(!t_i,p_i)_i\}_{i\in I}} = \{ \DT \mid \forall i\in I.~\stk{\DT\only i}\subseteq\RUN{!t_i}~\wedge~\pr{\DT\only i}\geq p_i~\}
\]
\end{definition}

To work with open terms, a \emph{substitution map} $\mu$ is a finite function from variables to terms, applied to a term as $\term{\mu M}$ which replaces every free variable $x$ in the domain of $\mu$ with the term $\term{\mu(x)}$. for a context $\Gamma=\term{x_1:t_1,..,x_n:t_n}$ the set $\SRUN{G}$ is a set of substitution maps $\mu$ that assigns each $x_i$ a runnable term $\term M\in\RUN{t_i}$, as follows.
\[
	\SRUN{G}~=~\{ \mu \mid \forall\term{x:t}\in\Gamma.~\term{\mu(x)}\in\RUN{t}\}
\]
We are now ready to prove the \emph{runnability lemma}, that typed terms are runnable.

% - - - - - - - - - - - - - - -
\begin{lemma}
\label{lem:run}
If $\term{G |- M:t}$ and $\mu\in\SRUN{G}$ then $\term{\mu M}\in\RUN{t}$.
\end{lemma}

\begin{proof}
See the Appendix.
\end{proof}
% - - - - - - - - - - - - - - -

The runnability lemma directly gives the main theorem of typed almost-sure termination.

% - - - - - - - - - - - - - - -
\begin{theorem}
If $\term{|- M: \e => d}$ then $\term M$ terminates with probability at least $\pr{\type{d}}$.
\end{theorem}

\begin{proof}
Let $\type{d}=\type{\{(!t_i,p_i)_i\}_{i\in I}}$. By Lemma~\ref{lem:run} we have $\Eval\e M\DT$ with $\DT\in\RUN{d}$, which means that $\pr{\DT\only i}\geq p_i$ for every $i\in I$, so that $\pr\DT \geq \pr{\type d}$.
\end{proof}
% - - - - - - - - - - - - - - -

%----------------------------------------------------------------------------------------------------
\section{Ground terms and finite Markov chains}

We have shown that the calculus introduced in this paper enjoys all the expected properties, in particular guaranteeing almost-sure termination. Several natural questions concerning the expressive power of the calculus remain open. Notably, it is of interest to investigate the behavior of the calculus under restrictions to low-order types, or when stronger termination properties than AST, such as \emph{positive} almost-sure termination (PAST), can be proved. In this section, we consider the fragment of \emph{ground terms}: those that involve only probabilistic branching and iteration, but not input and output. They are the terms of the following form.
\[
	\text{\emph{Gound terms:}}\qquad \term{M,N}~\coloneqq~\term i~\mid~\term{M;i->N}~\mid~\term{M^i}~\mid~\term{?a.a}
\]
We will show the following properties: one, typing of ground terms is decidable; two, ground terms readily encode as finite Markov chains; and three, finite Markov chains may be encoded as ground terms. These then have the following consequences. The last two properties imply that ground terms are equivalent to finite Markov chains. Since finite Markov chains are PAST (if they are AST), also ground terms are PAST. Finally, combining the first and last property, for the encoding of a Markov chain as a ground term, type inference calculates the probability of terminating with each final state.

As in the introduction, we may restrict types for ground terms to sub-distributions over choice labels, and abbreviate them by omitting input and return types.
\[
	\term{M : p_1i_1 * .. * p_ni_n}\quad\text{\emph{abbreviates}}\quad\term{M : \e => \e.p_1i_i * .. * \e.p_ni_n}
\]
Since ground terms are closed, the typing rules may further omit contexts. Thus restricted, they are as follows.
\[
       \infer{\term{i: 1i}}{}
\qquad \infer{\term{M;i->N : g ++ (p . d)}}{\term{M: g * pi} & \term{N: d}}
\qquad \infer{\term{M^i : /1{1-p} . d}}{\term{M: d * pi} & (p\neq 1)}
\qquad \infer{\term{?a.a : /12\top * /12\bot}}{}
\qquad \infer{\term{M : 0.d}}{}
\]

\begin{proposition}
\label{prop:ground type inference}
Typing for ground terms is decidable.
\end{proposition}

\begin{proof}
Immediate by induction on the term.
\end{proof}

We will use the following definition of Markov chains as probabilistic automata. We write $\DD X$ for the set of probability distributions over a set $X$, and write a distribution over $X=\{x_1,x_2,\dots\}$ as $\{p_1x_1,p_2x_2,\dots\}$.

\begin{definition}
A \emph{Markov chain}
\(C~=~(S,~s\in S,~F\subseteq S,~\delta:S\less F\to\DD{S})\)
is given by a set of \emph{states} $S$, a \emph{starting state} $s$, a set of \emph{final states} $F$, and a stochastic transition function $\delta$ from non-final states to states.
\end{definition}

The translation $\markov M$ from ground terms to Markov chains will take $\term M$ to a chain
\[
	\markov{M}~=~(S_M,~s_M,~F_M,~\delta_M)
\]
where internal states $S_M\less F_M$ represent subformulas and final states represent choices not captured by a later case construct. We consider chains modulo renaming of internal states, and when combining the translations of two terms we assume their internal states to be disjoint, but final states may overlap. To connect two chains, we use the notation $s\{t/i\}$ to rename a state, and the notation $\delta\{s/i\}$ which replaces all transitions to a (final) state $i$ with transitions to $s$.
\[
	\delta\{s/i\}(r)~=~\{~p(t\{s/i\}) \mid pt \in \delta(r)~\}
\qquad\qquad
	\begin{array}{r@{}l} i\{t/i\} &{}=t \\ s\{t/i\} &{}=s \quad (s\neq i) \end{array}
\]
The translation $\markov-$ is then as follows.
\[
\begin{aligned}
	\markov{i}      &= (\{s,i\},~s,~\{i\},~s\mapsto\D{i.1})
\\	\markov{M;i->N} &= (S_M\less\{i\}\cup S_N,~s_M,~F_M\less\{i\}\cup F_N,~\delta_M\{s_N/i\}\cup\delta_N)
\\  \markov{M^i}    &= (S_M\less\{i\},~s_M,~F_M\less\{i\},~\delta_M\{s_M/i\})
\\  \markov{?a.a}   &= (\{s,\top,\bot\},~s,~\{\top,\bot\},~s\mapsto\D{/12\top,/12\bot})
\end{aligned}
\]

\begin{example}
For the example term from the introduction, abbreviated as $\term{M^i}$ where $\term M$ is as below, the translation is as follows.
\[
	\term{M}~=~\term{?a.a ; \top->\heads ; \bot->(?b.b ; \top->\tails ; \bot -> i)}
\qquad\qquad
	\markov{M^i}~=~\vc{
\begin{tikzpicture}[thick,x=20pt,y=24pt,outer sep=0pt,inner sep=2pt]
	\node (r) at (1,2) {$r$};
	\node (s) at (2,1) {$s$};
	\node (u) at (0,1) {$u$};
	\node (v) at (1,0) {$v$};
	\node (w) at (3,0) {$w$};
	\node (h) at (0,0) {$\termcolor\heads$};
	\node (t) at (1,-1) {$\termcolor\tails$};

	\begin{scope}[->,>=stealth]
	
	\draw (1,2.7)--(r);
	\draw (r) --node[pos=0.6,anchor=south east]{$\scriptstyle{\frac12}$} (u);
	\draw (r) --node[pos=0.6,anchor=south west]{$\scriptstyle{\frac12}$} (s);
	\draw (s) --node[pos=0.6,anchor=south east]{$\scriptstyle{\frac12}$} (v);
	\draw (s) --node[pos=0.6,anchor=south west]{$\scriptstyle{\frac12}$} (w);
	\draw (u)--(h);
	\draw (v)--(t);
	
	\draw[rounded corners] (w)--(3.7,0.7)--(1.7,2.7)--(r);
	
	\end{scope}
\end{tikzpicture}}
\]
\[
\begin{array}{r@{}l}
	     \markov{?a.a}                            &{}= (\{r,\top,\bot\},~r,~\{\top,\bot\},~r\mapsto\D{/12\top,/12\bot})
\\[5pt]  \markov{?a.a ; \top->\heads}             &{}= (\{r,u,\heads,\bot\},~r,~\{\heads,\bot\},~r\mapsto\D{/12\bot,/12u}~u\mapsto\D{1\heads})
\\[5pt]  \markov{?b.b ; \top->\tails ; \bot -> i} &{}= (\{s,v,w,\tails,i\},~s,~\{\tails,i\}, s\mapsto\D{/12v,/12w}~v\mapsto\D{1\tails}~w\mapsto\D{1i}) \rule{30pt}{0pt}
\\[5pt]  \markov{M}                               &{}= (\{r,s,u,v,w,\heads,\tails,i\},~r,~\{\heads,\tails,i\},
\\[5pt]  \multicolumn{2}{r}{ r\mapsto\D{/12u,/12s}~s\mapsto\D{/12v,/12w}~u\mapsto\D{1\heads}~v\mapsto\D{1\tails}~w\mapsto\D{1i}~) }
\\[5pt]  \markov{M^i}                             &{}= (\{r,s,u,v,w,\heads,\tails\},~r,
\\[5pt]  \multicolumn{2}{r}{ r\mapsto\D{/12u,/12s}~s\mapsto\D{/12v,/12w}~u\mapsto\D{1\heads}~v\mapsto\D{1\tails}~w\mapsto\D{1r}~) }
\end{array}
\]
\end{example}


In the other direction, we give an interpretation $\sem C$ of a finite Markov chain as a ground term. The translation operates as follows, using states as choice labels. Given $C$ with starting state $s$ and transition $s\mapsto\{p_1t_1,\dots,p_nt_n\}\in\delta$, we remove this transition and consider the sub-chains $C(t_i)$ for $i\leq n$ with new starting state $t_i$, in which $s$ becomes a final state. These are translated inductively to $\term{M_i}$. Then $\sem C$ is given by a term $\term{M^s}$, iterating on $s$, where $\term M$ is a probabilistic sum that chooses $\term{M_i}$ with probability $p_i$. We formalize this below.

For $C=(S,s,F,\delta)$ and $t\in S\less\{s\}$ let $C(t)=(S,t,F\cup\{s\},\delta\less\{s\})$ where $\delta\less X$ is the transition function $\delta$ restricted to the domain $\mathrm{dom}(\delta)\less X$. Write $\term{p_1M_1+..+p_nM_n}$ for the term that chooses $\term{M_i}$ with probability $p_i$, given inductively below; recall that $\term{?a_p}$ represents an unfair coin toss, given either as primitive or encoded with $\term{?a}$ and iteration.
\[
	\term{1M}~=~\term{M} 
\qquad
    \term{pM+q_1N_1+..+q_nN_n}~=~\term{?a_p.a ; \top -> M; \bot -> (//{q_1}{1-p}N_1 + .. + //{q_n}{1-p}N_n) } 
\]
Then the interpretation of a Markov chain $C=(S,s,F,\delta)$ as a term is as follows.
\[
	\sem C~=~
	\left\{\begin{array}{ll}
		\term{s} & \text{if}~s~\in~F
	\\	\termcolor{(p_1\sem{C(t_1)}\term{+..+}p_n\sem{C(t_n)})^s} & \text{where}~\delta(s)=\D{p_1t_1,..,p_nt_n}	
	\end{array}\right.
\]



%For a distribution of terms
%\[
%	\DM~=~\D{p_1{\term{M_1}},..,p_n{\term{M_n}}}
%\]
%let the term $\term{\lfloor\DM\rfloor}$ be the internalization of that distribution:
%\[
%	\lfloor\D{1{\term{M}}}\rfloor~=~\term{M} 
%\qquad\qquad
%	\lfloor\D{p{\term{M}}}+\DM\rfloor~=~\term{?a_p.a ; \top-> M; \bot -> \lfloor{\fraction1{1-p}}\cdot\DM\rfloor} 
%\]
%Then the interpretation of a Markov chain $C$ as a term is as follows.
%\[
%	\sem C~=~
%	\left\{\begin{array}{ll}
%		\term{s} & \text{if}~s~\in~F
%	\\	{\termcolor\lfloor\D{{p_1\sem{C(t_1)}},..,p_n{\sem{C(t_n)}}}\rfloor^s} & \text{where}~\delta(s)=\D{p_1t_1,..,p_nt_n}	
%	\end{array}\right.
%\]

\begin{example}
Consider the following Markov chain $C$, where each arrow has probability $\frac13$.
\[
\vcenter{\hbox{
\begin{tikzpicture}[inner sep=2pt,outer sep=0pt,thick]
	\node[circle,draw] (a) at ( 90:1) {$a$} ;
	\node[circle,draw] (b) at (210:1) {$b$} ;
	\node[circle,draw] (c) at (330:1) {$c$} ;
	\node (i) at ( 90:2) {$i$} ;
	\node (j) at (210:2) {$j$} ;
	\node (k) at (330:2) {$k$} ;
	\begin{scope}[->,>=stealth]
	\draw (a.240) to[bend right] (b.60) ;
	\draw (b.0)   to[bend right] (c.180) ;
	\draw (c.120) to[bend right] (a.300) ;
	\draw (a.330) to[bend left]  (c.90) ;
	\draw (b.90)  to[bend left]  (a.210) ;
	\draw (c.210) to[bend left]  (b.330) ;
	\draw (a)--(i);
	\draw (b)--(j);
	\draw (c)--(k);
	\draw (120:2)--(a);
	\end{scope}
\end{tikzpicture}}}
\qquad
\begin{array}{l@{}l@{}l}
	    S      &{}~=~{}& \{a,b,c,i,j,k\}
\\[3pt]	s      &{}~=~{}& a
\\[3pt]	F      &{}~=~{}& \{i,j,k\}
\\[3pt]	\delta &{}~=~{}& a \mapsto \D{~/13b~,~/13c~,~/13i~}
\\[3pt]               && b \mapsto \D{~/13a~,~/13c~,~/13j~}
\\[3pt]               && c \mapsto \D{~/13a~,~/13b~,~/13k~}
\end{array}
\]
Its interpretation is as follows, writing $\term{M+N+P}$ for the term $\term{//13M+//13N+//13P}$.
\[
\begin{aligned}
	\sem C &~=~ \term{({\sem{C(b)}}+{\sem{C(c)}}+i)^a}
\\	       &~=~ \term{((a+{\sem{C(b)(c)}}+j)^b + (a+{\sem{C(c)(b)}}+k)^c+i)^a}
\\         &~=~ \term{((a+(a+b+k)^c+j)^b + (a+(a+c+j)^b+k)^c+i)^a}
\end{aligned}
\]
This term is typed as follows.
\[
\begin{aligned}
	    \term{(a+b+k)^c}         &: \type{/13a * /13b * /13k}
\\[5pt]	\term{ a+(a+b+k)^c+j}~~~ &: \type{/49a * /19b * /19k + /13j}
\\[5pt] \term{(a+(a+b+k)^c+j)^b} &: \type{/12a * /18k * /38j}
\\[5pt] \term{(a+(a+c+j)^b+k)^c} &: \type{/12a * /18j * /38k}
\\[5pt] \term{ (a+(a+b+k)^c+j)^b + (a+(a+c+j)^b+k)^c+i}~~~ &: \type{/13a * /16k * /16j * /13i}
\\[5pt] \term{((a+(a+b+k)^c+j)^b + (a+(a+c+j)^b+k)^c+i)^a} &: \type{/14k * /14j * /12i}
\end{aligned}
\]
The interpretation may be illustrated as follows, where the solid edges represent the tree-shape of the probabilistic sums, while the dashed edges represent the loops.
\[
\vc{
\begin{tikzpicture}[x=27pt,y=-27pt,inner sep=0.5pt,outer sep=0pt,thick]
	\node[circle,draw] (a) at (1,0) {$\term M$};
	\node[circle,draw] (b) at (0,1) {$\term N$};
	\node[circle,draw] (c) at (2,1) {$\term P$};
	\node[circle,draw] (d) at (0,2.4) {$\term Q$};
	\node[circle,draw] (e) at (2,2.4) {$\term R$};
	\node (i) at (1,-1)   {$\vphantom(i$};
	\node (j) at (-1,1)   {$~j~$};
	\node (k) at ( 3,1)   {$~k~$};
	\node (l) at (-1,2.4) {$~k~$};
	\node (m) at ( 3,2.4) {$~j~$};
	\begin{scope}[->,>=stealth]
	\draw (0,-1)--(a);
	\draw (a) to[bend right] (b);
	\draw (a) to[bend left]  (c);
	\draw (a) -- (i);
	\draw (b) -- (d);
	\draw (c) -- (e);
	\draw (b) -- (j);
	\draw (c) -- (k);
	\draw (d) -- (l);
	\draw (e) -- (m);
	\begin{scope}[dash pattern=on 3pt off 2pt]
	\draw (b.120) to[bend left]     node[pos=0.5,anchor=south east]{$\scriptstyle a~$} (a.150);
	\draw (c.60)  to[bend right]    node[pos=0.5,anchor=south west]{$~\scriptstyle a$} (a.30);
	\draw (d.120) to[bend left]     node[pos=0.5,anchor=south east]{$\scriptstyle b~$} (b.240);
	\draw (e.60)  to[bend right]    node[pos=0.5,anchor=south west]{$~\scriptstyle c$} (c.300);
	\draw (d.60)  to[bend left=10]  node[pos=0.5,anchor=west]      {$~\scriptstyle a$} (a.240);
	\draw (e.120) to[bend right=10] node[pos=0.5,anchor=east]      {$\scriptstyle a~$} (a.300);
	\end{scope}
	\end{scope}
\end{tikzpicture}
}
\qquad
\begin{array}{l@{}l}
        \term M &{}~=~\term{(N + P + i)^a}
\\[3pt] \term N &{}~=~\term{(a + Q + j)^b}
\\[3pt] \term P &{}~=~\term{(a + R + k)^c}
\\[3pt] \term Q &{}~=~\term{(a+b+k)^c}~=~\term{a+b+k}
\\[3pt] \term R &{}~=~\term{(a+c+j)^b}~=~\term{a+c+j}
\end{array}
\]
As the illustration suggests, the formalism may express any probabilistic tree with back-pointers. It may also express directed acyclic graphs, which yields a smaller interpretation of the Markov chain above, though still exponential, in the following term.
\[
	\term{ (b+c+i ; b -> a+c+j ; c -> (a+b+k ; b -> a+c+j)^c )^a }
\]
The types work out the same way.
\[
\begin{aligned}
	    \term{ b+c+i }~~~             &: \type{/13b * /13c * /13i}
\\[5pt] \term{ b+c+i ; b -> a+c+j}~~~ &: \type{/19a * /49c * /13i * /19j}
\\[5pt] \term{(a+b+k ; b -> a+c+j)^c} &: \type{/12a * /18j * /38k}
\\[5pt] \term{ b+c+i ; b -> a+c+j ; c -> (a+b+k ; b -> a+c+j)^c}     &: \type{/13a * /13i * /16j * /16k}
\\[5pt] \term{(b+c+i ; b -> a+c+j ; c -> (a+b+k ; b -> a+c+j)^c )^a} &: \type{/12i * /14j * /14k}
\end{aligned}
\]
\end{example}

\noindent
The translations $\markov-$ and $\sem-$ above demonstrate the following.

\begin{proposition}
Ground terms of the \PFMC\ are equivalent to finite Markov chains.
\end{proposition}

\noindent
Markov chains are PAST if they are AST, i.e.\ if any path may eventually reach a final state. For typed ground terms, which are AST, we then have the following.

\begin{proposition}
Typed ground terms of the \PFMCT\ are PAST.
\end{proposition}

\noindent
Combining type inference for ground terms (Proposition~\ref{prop:ground type inference}) and the translation $\sem-$, we have the following result.

\begin{proposition}
Encoding and typing a finite Markov chain computes its return distribution: if ${\termcolor\sem{C}}:\type{p_1`s_1*..*p_n`s_n}$ then $C$ terminates in the state $s_i$ with probability $p_i$.
\end{proposition}

\begin{proof}
By construction.
\end{proof}


%----------------------------------------------------------------------------------------------------
\section{Moving past PAST}

Further exploring the expressive power of the calculus, in this section we demonstrate that typed terms in general are AST, but not PAST: that is, we will give an example of a term with infinite expected running time, despite terminating with probability one. Our example relies crucially on higher-order functions.

We will use the default choice $\star$ as imperative \emph{skip}, and its case as sequential composition, $\term{M;N}=\term{M;\star ->N}$. Consider the following term, for a given $\term M$.
\[
	\term{[M].(\,(<f>.[f;f].i)\, + \,\star)^i }
\]
With probability $\fraction12$ it composes the term on the stack with itself and iterates, or it terminates. Evaluating this term on the machine gives the following return family, for the sequence of terms $\term{M_n}$ given by $\term{M_0}=\term M$ and $\term{M_{n+1}}=\term{M_n;M_n}$.
\[
	\{(\term{M_n},\star)_{\top^n\bot}\mid n\in\N\}~=~\{\,(\term M,\star)_{\bot}~,~(\term{M;M},\star)_{\top\bot}~,~(\term{(M;M);(M;M)},\star)_{\top\top\bot}~,~\dots~\}
\]
We may then evaluate the term $\term{M_n}$ left on the stack by post-composing with $\term{<g>.g}$, which pops and runs it, possibly using further base values further down on the stack. If $\term M$ terminates successfully with $\star$ in $m$ steps, $\term{M_n}$ takes $m\times 2^n$ steps. For example, assuming a natural numbers type $\type{\N}$ with primitives zero $\term{0:\N}$ and addition $\term{`+:\N\,\N => \N.1\star}$, we let $\term M$ be the successor function $\term{[1].`+: \N => \N.1\star}$. The term below left then gives the return family below right, taking $m\times 2^n$ machine steps (for some small constant $m$) to compute the number $\term{2^n:\N}$ from zero and successor. Since the probability of returning $\term{2^n}$ is $\fraction 1{2^{n+1}}$, the expected evaluation time on the machine is infinite: the term is not PAST.
\[
	\term{[0].[[1].`+].(\,(<f>.[f;f].i)\, + \,\star)^i ; <g>.g ~:~\e => \N . 1\star }
\qquad\qquad
	\{ (\term{2^n},\star)_{\top^n\bot}\mid n\in\N \}
\]
The term is typed: for any $\type{!t}$, if $\term M$ is typed $\term{M:!t => !t.1\star}$ then we have $\term{M^n : !t => !t.1\star}$, such as in the case above where $\type{!t}=\type\N$. For the loop we then have the following.
\[	
	\vc{\infer{\term{|- ((<f>.[f;f].i)\,+\,\star)^i: s => s.1\star}}{\term{|- (<f>.[f;f].i)\,+\,\star: s => s./12i ~*~ s./12\star }}} \qquad \text{where} \quad \type{s}~=~\type{!t => !t.1\star}
\]
The term $\term{<g>.g : (!t => !t.1\star)\,!t => !t.1\star}$ takes $\term{M^n : !t => !t.1\star}$ off the stack and evaluates it with an appropriate remaining stack of type $\type{!t}$. The existence of a typed term that is not PAST then demonstrates the following.

\begin{proposition}
The typed probabilistic FMC is not positive almost-surely terminating.
\end{proposition}

%----------------------------------------------------------------------------------------------------

\section{Conclusion}

We have demonstrated a probabilistic interpretation of the branching and iteration in the FMC to give a type system for almost-sure termination, the \PFMCT. We believe this to be a natural account of almost-sure termination for the following reasons.

The central idea, with which we started the introduction, is to take the familiar typed Conway iterator and make it probabilistic. Immediately, this gives an account of typed almost-sure termination. We explored the consequences of this idea within the \emph{control} extension of the FMC, consisting of the \emph{choice}, \emph{case}, and \emph{loop} constructs. In itself, this forms a minimal basis for imperative control flow, modelling conditionals, constants, case switches, exception handling, and iteration, all on simple abstract machine with a single continuation stack, both typed and untyped---unlike the Conway iterator, which must be typed. Combined with our probabilistic generator, the control constructs form the \emph{ground terms} of the \PFMC, which we showed to be equivalent to finite Markov chains, where typing computes the return distribution of the chain.

The full \PFMC\ is then an orthogonal and conservative combination of the control extension with the $\lambda$-calculus and the probabilistic generator. Each fragment operates on an independent stack of the extended Krivine machine, simply by pushing and popping. The type system is a conservative extension of simple types for the $\lambda$-calculus, where each fragment contributes a natural component: the control fragment, a formal sum of return values indexed by \emph{choice} labels; the generator, a probability (sub-)distribution over these choice labels. The combination of probabilistic almost-sure termination and higher-order functions then gives the calculus expressive power beyond positive AST.

In future work, we intend to further analyse the expressive power of the calculus and fragments of it. At present, being an extension of simple types, it is clear that expressive power is limited. Addressing this, we expect that the type system will generalize to dependent types and intersection types along standard lines. This is not idle speculation: the principle was already demonstrated for intersection types for higher-order store in the FMC~\cite{Heijltjes-2025-FSCD}.


%----------------------------------------------------------------------------------------------------

\bibliography{FMC-AST}

\newpage

\appendix

\section{Appendix}

\subsection*{Running the machine}

\begin{example}
\label{example:run}
Below we give a successful run of the machine for our running example term, with the definition of $\term{M+N}$ unfolded to its primitives, and the trace $\bot\,\bot\,\top$.
\[
	\term{M}~=~\term{(\heads + (\tails +\,i))^i}~=~\term{(?a.a ; \top->\heads ; \bot->(?b.b ; \top->\tails ; \bot -> i))^i}
\]
\[
\begin{array}{@{(\,}r@{\,,\,}l@{\,,\,}l@{\,,\,}r@{\,)}}
             \bot\,\bot\,\top & \e &         \term{(?a.a ; \top->\heads ; \bot->(?b.b ; \top->\tails ; \bot -> i))^i} & \e
\\\hline     \bot\,\bot\,\top & \e & \phantom(\term{?a.a ; \top->\heads ; \bot->(?b.b ; \top->\tails ; \bot -> i)} & \term{(i -> M)}
\\\hline     \bot\,\bot\,\top & \e & \phantom(\term{?a.a ; \top->\heads} & \term{(\bot->(?b.b ; \top->\tails ; \bot -> i))~(i -> M)}
\\\hline     \bot\,\bot\,\top & \e & \phantom(\term{?a.a} & \term{(\top->\heads)~(\bot->(?b.b ; \top->\tails ; \bot -> i))~(i -> M)}
\\\hline           \bot\,\top & \e & \phantom(\term{\bot} & \term{(\top->\heads)~(\bot->(?b.b ; \top->\tails ; \bot -> i))~(i -> M)}
\\\hline           \bot\,\top & \e & \phantom(\term{\bot} &                \term{(\bot->(?b.b ; \top->\tails ; \bot -> i))~(i -> M)}
\\\hline           \bot\,\top & \e & \phantom(\term{?b.b ; \top->\tails ; \bot -> i} &                               \term{(i -> M)}
\\\hline           \bot\,\top & \e & \phantom(\term{?b.b ; \top->\tails} &                               \term{(\bot -> i)~(i -> M)}
\\\hline           \bot\,\top & \e & \phantom(\term{?b.b} &                               \term{(\top->\tails)~(\bot -> i)~(i -> M)}
\\\hline                 \top & \e & \phantom(\term{\bot} &                               \term{(\top->\tails)~(\bot -> i)~(i -> M)}
\\\hline                 \top & \e & \phantom(\term{\bot} &                                              \term{(\bot -> i)~(i -> M)}
\\\hline                 \top & \e & \phantom(\term{i}    &                                                          \term{(i -> M)}
\\\hline                 \top & \e & \term{(?a.a ; \top->\heads ; \bot->(?b.b ; \top->\tails ; \bot -> i))^i} & \e
\\\hline                 \top & \e & \phantom(\term{?a.a ; \top->\heads ; \bot->(?b.b ; \top->\tails ; \bot -> i)} & \term{(i -> M)}
\\\hline                 \top & \e & \phantom(\term{?a.a ; \top->\heads} & \term{(\bot->(?b.b ; \top->\tails ; \bot -> i))~(i -> M)}
\\\hline                 \top & \e & \phantom(\term{?a.a} & \term{(\top->\heads)~(\bot->(?b.b ; \top->\tails ; \bot -> i))~(i -> M)}
\\\hline                   \e & \e & \phantom(\term{\top} & \term{(\top->\heads)~(\bot->(?b.b ; \top->\tails ; \bot -> i))~(i -> M)}
\\\hline                   \e & \e & \phantom(\term{\heads} &              \term{(\bot->(?b.b ; \top->\tails ; \bot -> i))~(i -> M)}
\\\hline                   \e & \e & \phantom(\term{\heads} &                                                        \term{(i -> M)}
\\\hline                   \e & \e & \phantom(\term{\heads} & \e
\end{array}
\]
\end{example}


\subsection*{Proving the Runnability Lemma}

Here we prove Lemma~\ref{lem:run}, the runnability lemma. Recall the statement of the Lemma:

\medskip

\noindent
\textbf{Runnability lemma:}\quad\textit{
If $\term{G |- M:t}$ and $\mu\in\SRUN{G}$ then $\term{\mu M}\in\RUN{t}$.
}

\begin{proof}
By induction on the typing derivation.
\begin{itemize}

	% Variable
	\item \emph{Variable:} 
\[
\infer{\term{G, x:t |- x:t}}{}
\]
By the definition of $\SRUN{G}$ we have $\term{\mu x}\in\RUN t$.

	% Choice
	\item \emph{Choice:}
\[
\infer{\term{G |- i: !s => !s.1i}}{}
\]
Let $S\in\RUN{!s}$ and note that $\term{\mu i}=\term i$ for any $\mu$. We have the evaluation $\eval\e Si Si$ for the empty trace. Then for the frontier $\{\e\}$ we have $\Eval Si{\{(S,i)_\e\}}$, with the singleton return family of $S$ and $i$ for the empty trace. Since $S\in\RUN{!s}$ and $p(\e)=1$ it follows that $\{(S,i)_\e\}\in\RUN{!s.1i}$.

	% Zero
	\item \emph{Zero}
\[
\infer{\term{G |- M : !s => 0 . d}}{}
\]
For any $S$ and $\mu$ we have $\Eval S{\mu M}\varnothing$. Since $\varnothing\subseteq\RUN{!t}$ for any $\type{!t}$ and $\pr{\varnothing\only i}\geq 0$ for any $i$ we have $\varnothing\in\RUN{0.d}$.

	% Expand
	\item \emph{Expand:}
\[
\infer{\term{G |- M: !s\,!r => !r\,d}}{\term{G |- M: !s => d}}
\]
Let $R\in\RUN{!r}$, $S\in\RUN{!s}$, and $\mu\in\SRUN{G}$. The inductive hypothesis gives the evaluation $\Eval SM\DT$. For each $(T,i)_t\in\DT$ we have the evaluation below left. By induction on the evaluation relation $\evalarrow$ we get the evaluation below right.
\[
\eval tS{\mu M}Ti
\qquad\qquad
\eval t{RS}{\mu M}{RT}i
\]
This gives $\Eval{RS}{\mu M}{R\,\DT}$. Since $R\in\RUN{!r}$ and $\stk{\DT\only i}\subseteq\RUN{!t_i}$, where $\type{!t_i}$ is the type for $i$ in $\type d$, we have $\stk{R\,\DT\only i} \subseteq \RUN{!r\,!t_i}$. Since also $\pr{R\,\DT\only i}=\pr{\DT\only i}\geq p_i$ we have $R\,\DT\in\RUN{!r\,d}$. 

	% Push
	\item \emph{Push:} 
\[
\infer{\term{G |- [N].M : !s => d}}{\term{G |- N:r} && \term{G |- M : r\,!s => d}}
\]
Let $S\in\RUN{!s}$ and $\mu\in\SRUN{G}$. By induction we have $\term{\mu N}\in\RUN{r}$ and $\term{\mu M}\in\RUN{r\,!s=>d}$, and for the latter the evaluation $\Eval {S\,\term{\mu N}}{\mu M}\DT$. Then for every $(T,i)_t\in \DT$ we have the evaluation below left, from which we construct the derivation below right.
\[
	\eval t{S\,\term{\mu N}}{\mu M}Ti
\qquad\qquad
	\vc{\infer[\Rpush]{ \eval tS{\mu([N].M)} Ti }{ \eval t{S\,\term{\mu N}}{\mu M} Ti } }
\]
Then $\Eval S{\mu([N].M)}\DT$ and $\term{\mu([N].M)}\in\RUN{!s => d}$.

	% Pop
	\item \emph{Pop:}
\[
\infer{\term{G |- <x>.M : r\,!s => d}}{\term{G , x:r |- M: !s => d}}
\]
Let $S\in\RUN{!s}$, $\term N\in\RUN r$, and $\mu\in\SRUN{G}$. Let $\mu'=\mu\{\term N/x\}$, the substitution map that sends $x$ to $\term N$ and every other variable $y$ to $\mu y$, so that $\mu'\in\SRUN{G,x:r}$. By induction we have $\Eval S{\mu'M}\DT$, given by the evaluation below for every $(T,i)_t\in\DT$, from which we construct the evaluation below right.
\[
	\eval tS{\mu'M} Ti
\qquad\qquad
	\vc{ \infer[\Rpop]{ \eval t{S\,\term{N}}{\mu(<x>.M)}Ti } {\eval tS{\mu'M}Ti} } 
\]
Then $\Eval{S\,\term N}{\mu(<x>.M)}\DT$ and $\term{\mu(<x>.M)}\in\RUN{r\,!s=>d}$.


	% Sample
	\item \emph{Sample:}
\[
\infer{\term{G |- ?a.M : !r => (/12 . g) ++ (/12 . d)}}{\term{G , a: \e => \e.1\top |- M: !r => g} && \term{G , a: \e => \e.1\bot |- M: !r => d}}
\]
Let $R\in\RUN{R}$ and $\mu\in\SRUN{G}$. Define $\mu_\top=\mu\{\term\top/a\}$ and $\mu_\bot=\mu\{\term\bot/a\}$ so that $\mu_\top\in\SRUN{G , a: \e => \e.1\top}$ and $\mu_\bot\in\SRUN{G , a: \e => \e.1\bot}$. By induction we have return families $\DS$ and $\DT$ where $\Eval R{\mu_\top M}\DS$ and $\Eval R{\mu_\bot M}\DT$, given by the following evaluations for every $(S,i)_s\in\DS$ and $(T,j)_t\in\DT$.
\[
\eval sR{\mu_\top M}Si \qquad\qquad  
\eval tR{\mu_\bot M}Tj
\]
For these we have the following derivations.
\[
\infer[\Rtop]{ \eval{\top\,s}R{\mu(?a.M)}Si }{ \eval sR{\mu_\top M}Si } \qquad
\infer[\Rbot]{ \eval{\bot\,t}R{\mu(?a.M)}Tj }{ \eval tR{\mu_\bot M}Tj }
\]
Then for $\DU=(\top\pfx\DS)\cup(\bot\pfx\DT)$ we have $\Eval S{\mu(?a.M)}\DU$. It remains to show that $\DU\in\RUN{/12.g ++ /12.d}$. For any indexed pair in $\type{/12.g ++ /12.d}$, by the definition of $(\type{++})$ we have three cases.
\begin{itemize}
	\item $\type{(!t,/12p+/12q))_i}$ where $i\in\floor{\type g}\cap\floor{\type d}$, $\type{(!t.p)_i}\in\type g$ and $\type{(!t.q)_i}\in\type d$. Then since $\stk{\DS\only i}\subseteq\RUN{!t}$ and $\stk{\DT\only i}\subseteq\RUN{!t}$ we have $\stk{\DU\only i}\subseteq\RUN{!t}$, and since $\pr{\DS\only i}\geq p$ and $\pr{\DT\only i}\geq q$ we have $\pr{\DU\only i}=\pr{\top\pfx\DS\only i}+\pr{\bot\pfx\DT\only i}\geq \frac 12p+\frac 12q$.
	\item $\type{(!t,/12p)_i}$ in $\type{/12.g-{\floor\delta}}$. Since $\stk{\DS\only i}\subseteq\RUN{!t}$ we have $\stk{\DU\only i}\subseteq\RUN{!t}$, and since $\pr{\DS\only i}\geq p$ we have $\pr{\DU\only i}=\pr{\top\pfx\DS\only i}\geq\frac 12p$.
	\item The case for $\type{/12.d-{\floor\gamma}}$ is analogous to the previous.
\end{itemize}

	% Sequence
	\item \emph{Sequence:} 
\[
\infer{\term{G |- M;i->N : !r => g ++ p.d}}
      {\term{G |- M: !r => g * !s.pi} && 
       \term{G |- N: !s => d}}
\]
Let $R\in\RUN{!r}$ and $\mu\in\SRUN{G}$. By induction we have $\term{\mu M}\in\RUN{!r => g * !s.pi}$ and $\Eval R{\mu M}\DS$. Let this be given by an evaluation for each $(S,j)_s\in\DS$ of the form below left. For each $(S,j)_s$ where $j=i$ we have $S\in\RUN{!s}$, and by induction on $\term{N}$ a return family $\DT_s$ such that $\Eval S{\mu N}{\DT_s}$. Let this be given by the evaluation below right for each $(T,k)_t\in\DT_s$.
\[
	\eval sR{\mu M}Sj
\qquad
	\eval tS{\mu N}Tk
\] 
For each $(S,j)_s$, if $j\neq i$  we construct the derivation below left, and if $j=i$ then for each $(T,k)_t\in\DT_s$ we construct that below right. 
\[
	\infer[\Rrej]{ \eval sR{\mu(M;i->N)}Sj }{ \eval sR{\mu M}Sj }
\qquad\qquad
	\infer[\Rsel]{ \eval{st}R{\mu(M;i->N)}Tk }{ \eval sR{\mu M}Sj && \eval tS{\mu N}Tk }
\]
Then we have $\Eval R{\mu(M;i->N)}\DT$ where $\DT$ is given as follows.
\[
	\DT~=~(\DS\less i) \cup \bigcup_{s\in\floor{\DS\only i}} s\pfx\DT_s
\]
It remains to show that $\DT\in\RUN{g ++ p.d}$.

\begin{itemize}
	\item
First, for the return types, for all $j\in\floor{\type g}\cup\floor{\type d}$ we need that $\stk{\DT\only j}\subseteq\RUN{!t}$ where $\type{!t}$ is the type for $j$ in $\type g$ and $\type d$ (the definition of $(\type{++})$ requires that they agree). By definition any indexed pair in $\DT$ is either $(S,j)_s\in\DS\less i$ or $(T,k)_t\in\DT_s$. The former case follows since $\DS\in\RUN{g}$ and so $\stk{\DS\only j}\subseteq\RUN{!t}$ and $S\in\subseteq\RUN{!t}$. The latter case follows since $\DT_s\in\RUN{d}$, so that $\stk{\DS\only j}\subseteq\RUN{!t}$ and $T\in\RUN{!t}$.
	\item
Second, for the probabilities, for all $j\in\floor{\type g}\cup\floor{\type d}$ it must be shown that $\pr{\DT\only j}\geq p_j + pq_j$ where $p_j$ is the probability of $j$ in $\type g$ and $q_j$ that in $\type d$. We compute as follows.
\[
\begin{aligned}
\pr{\DT\only j} 
    &~=~    \pr{((\DS\less i)~\cup~\bigcup_{s\,\in\,\floor{\DS\only i}} s\pfx\DT_s)\only j}
\\  &~=~    \pr{(\DS\less i)\only j} + \sum_{s\in\floor{\DS\only i}} (\pr{s}\times\pr{\DT_s\only j})
\\  &~\geq~ p_j + \sum_{s\in\floor{\DS\only i}} (\pr{s}\times q_j)
\\  &~\geq~ p_j + pq_j
\end{aligned}
\]
For the third step (the first inequality), since $\DS\less i\in\RUN{g}$ we have $\pr{\DS\only j}\geq p_j$ for all $j\in\floor{\type g}$. For the last step, since $\DT_s\in\RUN{d}$ for all $s$ we have $\pr{\DT_s\only j}\geq q_j$ for all $j\in\floor{\type d}$. Then since $\DS\only i\in\RUN{!s.pi}$ we have $\pr{\DS\only i}\geq p$.
\end{itemize}

It follows that $\DT\in\RUN{g ++ p.d}$ and hence $\term{\mu(M;i->N)}\in\RUN{!r => g ++ p.d}$.


	% Iterate
	\item \emph{Iterate:}
\[
\infer{\term{G |- M^i: !s => /1{1-p} . d}}{\term{G |- M: !s => d + !s.pi} && (p\neq 1)}
\]
Let $S\in\RUN{!s}$ and $\mu\in\SRUN{G}$. We construct a series of return families $\DT_n$ for all $n\in\N^+$ by induction on $n$, and show the following.
\[
	\DT_n\in\RUN{(q_n.d) + !s.p^ni} \quad\text{where}\quad q_n=1+p+p^2+p^3+\dots+p^{n-1}
\]

\begin{itemize}
\item
The case $\DT_1$ is given by induction on $\term M$ by $\Eval S{\mu M}{\DT_1}$. 

\item
For the case $\DT_{n+1}$, for each $(R,i)_r\in\DT_n$, by induction on $\term M$ let $\Eval R{\mu M}{\DT_r}$ and define $\DT_{n+1}$ as follows.
\[
	\DT_{n+1}~=~(\DT_n\less i) \cup \bigcup_{r\in\floor{\DT_n\only i}} r\pfx\DT_r
\]
By induction we have $\DT_n\in\RUN{q_n.d + !s.p^ni}$. By similar reasoning to the previous \emph{sequencing} case we have the following.
\[
	\DT_{n+1}\in\RUN{q_n.d + p^n.(d + !s.pi)}~=~\RUN{q_{n+1}.d + !s.p^{n+1}i}
\]
\end{itemize}

Observe that by definition $\DT_n\less i~\subseteq~\DT_{n+1}\less i$. Then we define $\DT$ as the following supremum.
\[
	\DT~=~\sup_{n\in\N^+}\DT_n\less i
\]
It remains to show that $\Eval S{\mu(M^i)}\DT$ and $\DT\in\RUN{/1{1-p}.d}$.

\begin{itemize}
	\item
If $(T,j)_t\in\DT$, then by definition $(T,j)_t\in\DT_n\less i$ for some $n$. Assume this is the least such $n$. Then by induction, going backwards from $n$, we give a sequence of stacks $S_m$ and traces $s_m$ and $t_m$ for all $m<n$ such that $S_0=S$, $s_0=\e$, $s_m\,t_m=t$, and $\eval {t_m}{S_m}{\mu(M^i)} Tj$, and $(S_m,i)_{s_m}\in\DT_m$ for $m\neq 0$.

\begin{itemize}
	\item
For the case $m=n-1$, since $(T,j)_t\in\DT_n$, by definition it is either in $\DT_{n-1}\less i$, which is ruled out by assumption, or it is given by an evaluation $\Eval R{\mu M}{\DT_r}$ with $R=S$ if $n=1$, and otherwise $(R,i)_s\in\DT_{n-1}$ and $t=sr$, with the evaluation below left. Let $S_{n-1}=R$, $s_{n-1}=s$, and $t_{n-1}=r$. Since $i\neq j$, this gives the derivation below right.
\[
\eval rR{\mu M} Tj \qquad \infer[\Rexit]{ \eval {t_{n-1}}{S_{n-1}}{\mu(M^i)} Tj }{ \eval {t_{n-1}}{S_{n-1}}{\mu M} Tj }
\]

	\item
For the cases $0<m<n$, by induction let $(S_m,i)_{s_m}\in\DT_m$ with $t=s_mt_m$ and the evaluation below left. By the definition of $\DT_m$, $(S_m,i)_{s_m}$ is given by an evaluation $\Eval R{\mu M}{\DT_r}$ with $R=S$ if $m=1$ and otherwise $(R,i)_s\in\DT_{m-1}$ with $s_m=sr$, giving the evaluation below right.
\[
	\eval{t_m}{S_m}{\mu(M^i)}Tj
\qquad\qquad
	\eval rR{\mu M}{S_m}i
\] 
To show the statement for $m-1$, let $S_{m-1}=R$, $s_{m-1}=s$, and $t_m=r\,t_{m-1}$, and construct the derivation below.
\[
	\infer[\Rloop]{ \eval {t_{m-1}}{S_{m-1}}{\mu(M^i)} Tj }
	{ \eval {s_{m-1}}{S_{m-1}}{\mu M} {S_m}i && \eval {t_m}{S_m}{\mu(M^i)} Tj }
\]
\end{itemize}
It follows that for every $(T,j)_t\in\DT$ there is an evaluation $\eval tS{\mu(M^i)}{T_t}{j_t}$, and hence $\Eval S{\mu(M^i)}\DT$.

	\item
For all $n\in\N^+$ we have $\DT_n\in\RUN{q_n.d + !s.p^ni}$ and hence $\DT_n\less i\in\RUN{q_n.d}$. Since each $T$ in $\stk{\DT\only j}$ is in some $\stk{\DT\only j}$ we have $T\in\RUN{!t}$ where $\type{!t}$ is the stack for $j$ in $\type d$. Next, for $j\neq i$ we have $\pr{\DT_n\only j} \geq q_np_j$ where $q_n=1+p+p^2+\dots +p^{n-1}$ and $p_j$ is the probability for $j$ in $\type{d}$. This gives the following. 
\[
	\pr{\DT\only j} \geq (1+p+p^2+p^3\dots)\times p_j ~=~ \frac1{1-p}\,p_j 
\]
\end{itemize}

It follows that $\DT\in\RUN{/1{1-p}.d}$ and hence $\term{\mu(M^i)}\in\RUN{!s => /1{1-p}.d}$.
\end{itemize}
\end{proof}

\end{document}


%====================================================================================================




The following term $\term Z$ generates a geometric series of Church numerals.
\[
	\term{Z}~=~\term{ <f>.<x>.[x].~(<n>.~([n].*~+~[[n].f].i))^i ; <m>.m }~\sim~\D{\cno./12,~\cni./14,~\cnii./18,~\dots}
\]
Let $\term M=\term{<n>.([n].*~+~[[n].f].i)}$ so that the above term is $\term{ <f>.<x>.[x].M^i ; <m>.m }$. For a given $\term N$, the term$\term{[N].M^i}$ evaluates as follows.
\[
\begin{aligned}
\term{[N].M^i}  
 ~ \rws ~ & \term{[N].M;i->M^i}
\\    = ~ & \term{[N].<n>.([n].*~+~[[n].f].i);i->M^i}
\\  \rw ~ & \term{([N].*~+~[[N].f].i);i->M^i}
\\ \sim ~ & \term{[N].* ~+~ [[N].f].M^i}
\end{aligned}
\]
The overall term $\term{ <f>.M^i ; <x>.x }$ then generates the geometric series as follows.
\[
\begin{array}{@{}r@{}l@{}}
	       & \term{ <f>.<x>.[x].M^i ; <m>.m }
\\  ~\rws~ & \term{ <f>.<x>.([x].* ; [[x].f].M^i) ; <m>.m }
\\ 	~\sim~ & \term{ <f>.<x>.[x].<m>.m ~+~ (<f>.<x>.[[x].f].M^i ; <m>;m) }
\\  ~\rw ~ & \term{ \cno ~+~ (<f>.<x>.[[x].f].M^i ; <m>;m) }
\\  ~\rws~ & \term{ \cno ~+~ (\cni ~+~ (<f>.<x>.[[[x].f].f].M^i ; <m>;m)) }
\end{array}
\]
Church numerals allow exponentiation by $\underline{m^n}=\cnn\,\cnm$. The term $\term{[\cnii].Z}$ then generates the following series.
\[
	\term{[\cnii].Z}~\sim~\D{\cni./12~,~\cnii./14~,~\cniv./18~,~..}
\]
Since Church numerals are in unary notation, the terms in the series grow exponentially. Stochastically generating a Church numeral with $\term{[\cnii].Z}$ is thus almost-surely terminating, but not \emph{positively} so.

There are two minor caveats. First, the above computations use reduction $\rw$ and equivalence $\sim$, but not evaluation on the machine. Second, since the calculus encodes constants as choices, it only admits finite datatypes. One may address both issues by introducing a natural numbers type $\type\N$ with primitive operations $\term{`+:\N\,\N=>\N}$ etc. Then the following term evaluates on the machine to return a single natural number according to the above series. (This has been confirmed experimentally.)
\[
	\term{[[0].*].[<n>.n ; [1].`+].[\cnii].Z}
\]
Note that Church numerals $\term{\cnn}$ follow a call--by--name semantics, while primitive operations on natural numbers have a call--by--value semantics. This is reflected in the base value $\term{[0].*}$ and successor function $\term{<n>.n ; [1].`+}$, which incorporate the encoding of call--by--value into call--by--name. 

It remains to show that the above terms can be typed. First, for any type $\type{s=>t}$, and indeed any type $\type{!s=>!t_I}$, the Church numerals and the term $\term Z$ can be typed as follows.
\[
\begin{aligned}
	    \term{ n }                &: \type{s=>t}
\\[5pt] \term{ f }                &: \type{(s=>t)\,s => t}
\\[5pt] \term{ [[n].f].i }        &: \type{\e => (s=>t).i}
\\[5pt] \term{ [n].*~+~[[n].f].i} &: \type{\e => (s=>t)./12* + (s=>t)./12i}
\\[5pt] \term{M}~=~\term{<n>.([n].*~+~[[n].f].i)} &: \type{(s=>t) => (s=>t)./12* + (s=>t)./12i}
\\[5pt] \term{M^i}                &: \type{(s=>t) => (s=>t)}
\\[5pt] \term{Z}~=~\term{ <f>.<x>.[x].~M^i ; <m>.m } &: \type{((s=>t)\,s => t)~(s=>t)~s~=>~t}
\end{aligned}
\]
It follows that $\term{[\cnii].Z}$ can be typed with the same type. The types of the natural number constructions are as follows.
\[
	\term{[0].* : \e=>\N} \qquad
\qquad
	\term{<n>.n ; [1]. `+ : (\e=>\N)=>\N}
\]
This gives the following overall type. 
\[
	\term{[[0].*].[<n>.n ; [1].`+].[\cnii].Z : \e=>\N}
\]
The Church numeral types $\type{((s=>t)\,s => t)~(s=>t)~s=>t}$ of $\term{\cnii}$ and $\term{Z}$ here are specialised as follows: that for $\term{\cnii}$ by replacing $\type{s=>t}$ with $\type{\e=>\N}$, and that for $\term Z$ by replacing $\type{s=>t}$ with $\type{(\e=>\N)=>\N}$.
\[
\begin{aligned}
	    \term{\cnii} &: \type{((\e=>\N)=>\N)\,(\e=>\N)=>\N}
\\[5pt]	\term{Z}     &: \type{(((\e=>\N)=>\N)\,(\e=>\N)=>\N)~((\e=>\N)=>\N)~(\e=>\N)~=>~\N}
\end{aligned}
\]
The existence of this typed term gives the following.

\begin{proposition}
The typed probabilistic FMC expresses terms that are not positively almost-sure terminating.
\end{proposition}
